---
title: "Simulation Studies Pipeline"
author: "Laurens Bogaardt"
date: "2023-06-07"
output:
  html_document:
    keep_md: true
    df_print: paged
    fig_width: 10
    fig.align: center
    toc: true
    toc_depth: 4
    toc_float: true
    theme: united
    code_folding: show
---

# Introduction

This document describes the pipeline to create ...

The document not only describes each step, it also runs all scripts in the correct order. This is done by knitting the documents using the packages _xfun_ and _rmarkdown_.

```{r results = FALSE, message = FALSE, warning = FALSE}
library(xfun)
library(rmarkdown)
library(tidyverse)
library(stringr)
Rscript_call <- function(fun, args) {} # Temporary trick to disable knitting.
```

# Analyse Pipeline

Next, all documents inside the pipelines are listed.

```{r}
rmarkdown.documents <- tibble(
  file.name = list.files(
    pattern = "\\.Rmd?$"
  )
) %>%
  filter(file.name != "Simulation-Studies-Pipeline.Rmd") %>%
  group_by(file.name) %>%
  mutate(
    source.code = paste(readLines(file.name, warn = FALSE), collapse = "\n")
  ) %>%
  ungroup()
rmarkdown.documents
```

## Check Packages

```{r}
packages.per.file <- rmarkdown.documents %>%
  group_by(file.name) %>%
  reframe(packages = gsub("library\\(\"?([^\\)]*?)\"?\\)", "\\1", str_extract_all(source.code, "library\\(\"?([^\\)]*?)\"?\\)")[[1]]))
packages.per.file
```

The documents are analysed to see what packages are required.

```{r}
required.packages <- packages.per.file %>%
  pull(packages) %>%
  unique() %>%
  sort()
required.packages
```

A check is done to see if all required packages are installed.

```{r}
missing.packages <- required.packages[sapply(required.packages, function(x){length(find.package(x, quiet = TRUE)) == 0})]
if (length(missing.packages) > 0) {
  cat(paste0("You are currently missing some packages. Please install these using:\ninstall.packages(c(\"", paste(missing.packages, collapse = "\", \""), "\"))"))
} else {
  cat("All required packages are installed.")
}
```

## Pipeline Sections

When applying the regular expression to the file `Simulation Studies Pipeline.Rmd`, the section title, type and content can also be obtained.

```{r}
pipeline.sections.regex <- "# ([^\\n]+?)\\n\\n([^#]*(##[^#]*)*)"
pipeline.sections <- rmarkdown.documents %>%
  filter(file.name == "Simulation Studies Pipeline.Rmd") %>%
  reframe(
    section.code = unlist(str_extract_all(source.code, pipeline.sections.regex))
  ) %>%
  mutate(
    section.title = str_replace_all(section.code, pipeline.sections.regex, "\\1"),
    section.content = str_replace_all(section.code, pipeline.sections.regex, "\\2")
  ) %>%
  filter(!section.title %in% c("Introduction", "Check Packages"))
pipeline.sections
```

We examine the document sections using the following regular expression.

```{r}
pipeline.documents.regex <- "## ([^\\n]+?)\\n\\nThe document \\[_(.*?)_\\]\\((.*?)\\) is knitted to the file \\[_(.*?)_\\]\\((.*?)\\)\\..*"
```

```{r}
pipeline.documents <- pipeline.sections %>%
  group_by(section.title) %>%
  reframe(
    subsection.code = unlist(str_extract_all(section.content, pipeline.documents.regex))
  ) %>%
  mutate(
    subsection.title = str_replace_all(subsection.code, pipeline.documents.regex, "\\1"),
    document.link.name = str_replace_all(subsection.code, pipeline.documents.regex, "\\2"),
    document.file.path = str_replace_all(subsection.code, pipeline.documents.regex, "\\3"),
    knit.link.name = str_replace_all(subsection.code, pipeline.documents.regex, "\\4"),
    knit.file.path = str_replace_all(subsection.code, pipeline.documents.regex, "\\5")
  )
pipeline.documents
```

## Document Structure

The following check shows which section title and document link name do not correspond, if any.

```{r}
pipeline.documents %>%
  select(section.title, subsection.title, document.link.name) %>%
  filter(subsection.title != document.link.name)
```

The following check shows which document link name and document file name do not correspond, if any.

```{r}
pipeline.documents %>%
  select(section.title, subsection.title, document.link.name, document.file.path) %>%
  filter(document.link.name != document.file.path)
```

The following check shows which referenced document file does not exist, if any.

```{r}
pipeline.documents %>%
  select(section.title, subsection.title, document.file.path) %>%
  filter(!file.exists(document.file.path))
```

The following check shows which document file path and knitted document file path do not correspond, if any.

```{r}
pipeline.documents %>%
  select(section.title, subsection.title, document.file.path, knit.file.path) %>%
  filter(gsub(".Rmd", ".html", gsub("\\s", "-", document.file.path)) != knit.file.path)
```

The following check shows which knitted document link name and knitted document file name do not correspond, if any.

```{r}
pipeline.documents %>%
  select(section.title, subsection.title, knit.link.name, knit.file.path) %>%
  filter(knit.link.name != knit.file.path)
```

## Included Documents

```{r}
pipeline.documents %>%
  select(file.name = subsection.title) %>%
  setdiff(
    rmarkdown.documents %>%
      filter(file.name != "Simulation Studies Pipeline.Rmd") %>%
      select(file.name)
  )
```

```{r}
rmarkdown.documents %>%
  filter(file.name != "Simulation Studies Pipeline.Rmd") %>%
  select(file.name) %>%
  setdiff(
    pipeline.documents %>%
      select(file.name = subsection.title)
  )
```

# Linear Regression Models

The _ABC_ ...

## Linear Regression Simulation Study.Rmd

The document [_Linear Regression Simulation Study.Rmd_](Linear Regression Simulation Study.Rmd) is knitted to the file [_Linear-Regression-Simulation-Study.html_](Linear-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Linear Regression Simulation Study.Rmd",
      output_format = "all",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
Rscript_call <- function(fun, args) {} # Temporary trick to disable knitting.
```

# Mixed Effects Models 

The _ABC_ ...

## Mixed Effects Model Simulation Study.Rmd

The document [_Mixed Effects Model Simulation Study.Rmd_](Mixed Effects Model Simulation Study.Rmd) is knitted to the file [_Mixed-Effects-Model-Simulation-Study.html_](Mixed-Effects-Model-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Mixed Effects Model Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## B-Spline Mixed Effects Model Simulation Study.Rmd

The document [_B-Spline Mixed Effects Model Simulation Study.Rmd_](B-Spline Mixed Effects Model Simulation Study.Rmd) is knitted to the file [_B-Spline-Mixed-Effects-Model-Simulation-Study.html_](B-Spline-Mixed-Effects-Model-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "B-Spline Mixed Effects Model Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## Multivariate Mixed Effects Model Simulation Study.Rmd

The document [_Multivariate Mixed Effects Model Simulation Study.Rmd_](Multivariate Mixed Effects Model Simulation Study.Rmd) is knitted to the file [_Multivariate-Mixed-Effects-Model-Simulation-Study.html_](Multivariate-Mixed-Effects-Model-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Multivariate Mixed Effects Model Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

# Binary Classification Models

The _DEF_ ...

## Logit Regression Simulation Study.Rmd

The document [_Logit Regression Simulation Study.Rmd_](Logit Regression Simulation Study.Rmd) is knitted to the file [_Logit-Regression-Simulation-Study.html_](Logit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Logit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## Probit Regression Simulation Study.Rmd

The document [_Probit Regression Simulation Study.Rmd_](Probit Regression Simulation Study.Rmd) is knitted to the file [_Probit-Regression-Simulation-Study.html_](Probit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Probit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## Gompit Regression Simulation Study.Rmd

The document [_Gompit Regression Simulation Study.Rmd_](Gompit Regression Simulation Study.Rmd) is knitted to the file [_Gompit-Regression-Simulation-Study.html_](Gompit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Gompit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

# Ordinal Classification Models

The _DEF_ ...

## Ordinal Logit Regression Simulation Study.Rmd

The document [_Ordinal Logit Regression Simulation Study.Rmd_](Ordinal Logit Regression Simulation Study.Rmd) is knitted to the file [_Ordinal-Logit-Regression-Simulation-Study.html_](Ordinal-Logit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Ordinal Logit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## Ordinal Probit Regression Simulation Study.Rmd

The document [_Ordinal Probit Regression Simulation Study.Rmd_](Ordinal Probit Regression Simulation Study.Rmd) is knitted to the file [_Ordinal-Probit-Regression-Simulation-Study.html_](Ordinal-Probit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Ordinal Probit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

## Ordinal Gompit Regression Simulation Study.Rmd

The document [_Ordinal Gompit Regression Simulation Study.Rmd_](Ordinal Gompit Regression Simulation Study.Rmd) is knitted to the file [_Ordinal-Gompit-Regression-Simulation-Study.html_](Ordinal-Gompit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Ordinal Gompit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```

# Multinomial Classification Models

The _DEF_ ...

## Multinomial Logit Regression Simulation Study.Rmd

The document [_Multinomial Logit Regression Simulation Study.Rmd_](Multinomial Logit Regression Simulation Study.Rmd) is knitted to the file [_Multinomial-Logit-Regression-Simulation-Study.html_](Multinomial-Logit-Regression-Simulation-Study.html).

```{r, message = FALSE, warning = FALSE}
system.time(
  Rscript_call(
    fun = render,
    args = list(
      input = "Multinomial Logit Regression Simulation Study.Rmd",
      output_dir = ".",
      intermediates_dir = "."
    )
  )
)
```
